---
title: "Assignment 2: Physical Properties of Lakes"
author: "Sasha Keller"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on the physical properties of lakes. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After completing your assignment, fill out the assignment completion survey in Sakai. 

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-01-25

## Setup

1. Verify your working directory is set to the R project file, 
2. Load the tidyverse, lubridate, and rLakeAnalyzer packages
3. Import the NTL-LTER physical lake dataset and set the date column to the date format. 
4. Using the `mutate` function, add a column called Month. Remove temperature NAs.
5. Set your ggplot theme (can be theme_classic or something else)

```{r}
getwd()

library(tidyverse)
library(lubridate)
#install.packages("rLakeAnalyzer")
library(rLakeAnalyzer)

NTLdata <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
NTLdata<- NTLdata %>%
  dropna(temperature_C)
NTLdata$sampledate<-as.Date(NTLdata$sampledate, format = "%m/%d/%y")

NTLdata<-mutate(NTLdata, Month = format(NTLdata$sampledate, "%m"))
view(NTLdata)
theme_set(theme_dark())
```

## Creating and analyzing lake temperature profiles

5. For the year 1993, plot temperature and dissolved oxygen profiles for all six lakes in the dataset (as two separate ggplots). Use the `facet_wrap` function to plot each lake as a separate panel in the plot. Plot day of year as your color aesthetic and use a reverse y scale to represent depth. 

What seasonal trends do you observe, and do these manifest differently in each lake?

In 4 of 6 lakes we see homogenization of temperature across th water column in Winter, indicating a frozen top. It appears they all stratify over the course of the year. There appear to be several straight lines in upper water column for all lakes throughout the year. It is unclear to me whether they have become non-stratified, indicating that these might be polymictic lakes, or if there is, perhaps, some lack of sensitivity in measuring equipment or other phenomenon at play. Central long lake does not appear to become non-stratified and is clearly a shallow polymictic lake. Tuesday lake lacks the winter lack of stratification of the other lakes, but does also appear tohave some of these non-stratified moments throughout the year.

```{r}
NTL1993Temp<- ggplot(subset(NTLdata, year4 == 1993),
                 aes(x=temperature_C, y=depth, color=daynum))+ 
  geom_point()+
  scale_y_reverse()+
  labs(x= expression("Temperature" (degree*C)), y= "Depth (m)", color="DOY")+
  facet_wrap(~ lakename)
NTL1993Temp

NTL1993O2<- ggplot(subset(NTLdata, year4 == 1993),
                 aes(x=dissolvedOxygen, y=depth, color=daynum))+ 
  geom_point()+
  scale_y_reverse()+
  labs(x= "Dissolved Oxygen", y= "Depth (m)", color="DOY")+
  facet_wrap(~ lakename)
NTL1993O2
```

6. Create a new dataset that calculates thermocline depths for all lakes on all dates (hint: you will need group by lake, year, month, DOY, and sample date).
7. Plot thermocline depth by day of year for your newly made dataset. Color each point by lake name, make the points 50% transparent, and choose a color palette other than the ggplot default. 
8. Create a boxplot of thermocline depth distributions split up by lake name on the x axis and by month as the fill color (hint: you will need to set Month as a factor). Choose a color palette other than the ggplot default, relabel axes and legend, and place the legend on the top of the graph.

Do all the lakes have a similar seasonal progression of thermocline deepening? Which lakes have the deepest thermoclines, and how does this relate to their maximum depth? 
They all appear to to obtain deeper thermoclines as the year progresses. Some of the lakes only have one thermocline measurement, which makes it difficult to interpret. It appears that thermoclines are absent in shallow lakes and increasingly deep with increasing lake depth.


```{r}
thermoplot<- NTLdata %>%
  group_by(lakename,daynum,Month,year4,sampledate) %>%
  summarise(thermocline = thermo.depth(wtr= temperature_C,depths = depth, seasonal = FALSE)) %>%
  filter(thermocline>1)

ggplot(thermoplot, aes(x= daynum, y=thermocline, color=lakename))+
  geom_point(alpha=0.5)+
  scale_color_brewer(palette="Accent")

ggplot(thermoplot,aes(x=lakename,y=thermocline, fill=as.factor(Month)))+
  geom_boxplot()+
  scale_color_brewer(palette="Accent")+
  labs(x="Lake Names", y="Thermocline Depth", color= "Month")+
  theme_dark()+
  theme(legend.position="top")
```

