---
title: "Assignment 8: Time Series Analysis"
author: "Sasha Keller"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on time series analysis

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, check your PDF against the key and then submit your assignment completion survey at https://forms.gle/dKEutwXiFewkSTwN9

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-03-29

## Setup

1. Verify your working directory is set to the R project file. Load the tidyverse, lubridate, trend, forecast, and dataRetrieval packages. Set your ggplot theme (can be theme_classic or something else).


```{r setup}
getwd()
library(tidyverse)
library(lubridate)
#install.packages("trend")
library(trend)
#install.packages("forecast")
library(forecast)
library(dataRetrieval)

theme_set(theme_dark())

```

## Data Import and Processing

2. Import discharge data (parameter 00060) from Clear Creek, Colorado (site 06719505) from the start of 1975 through the end of 2021.
3. Graph Clear Creek discharge over time. 
4. Create a new data frame with the sum of monthly discharge in acre-feet per month.

```{r}
CCDischarge<- readNWISdv(siteNumbers = "06719505",parameterCd ="00060",startDate = "1975-1-1",endDate = "2021-12-31")
names(CCDischarge)[4:5] <- c("Discharge", "Approval.Code")

CCDisPlot<-CCDischarge %>%
  ggplot(aes(x=Date,y=Discharge))+
  geom_line()
CCDisPlot

#Monthly Processing  
Months <- data.frame(Date_monthrounded = seq.Date(from = as.Date("1975-1-1"), to = as.Date("2021-12-31"), by = "month"))

CCDischarge_monthly <- CCDischarge %>%
  mutate(Year = year(Date),
         Month = month(Date)) %>%
  group_by(Year, Month) %>%
  summarise(Discharge_acftmo = sum(Discharge)*1.98347)
```


## Time Series Decomposition
5. Create a time series of discharge from the monthly data frame. Make sure to add start and end dates like we did in class.
6. Decompose the time series using the `stl` function.
7. Visualize the decomposed time series.

```{r}
CC_ts<-ts(CCDischarge_monthly$Discharge_acftmo,frequency=12)

CC_Decomposed <- stl(CC_ts, s.window = "periodic")

plot(CC_Decomposed)
```

8. How do the seasonal and trend components of the decomposition compare to the Neuse River discharge dataset? 

> Seasonal: The Clear Creek seasonal seems to be a larger component of the overall discharge than the Neuse.

> Trend: While the Neuse Trend line seems to have consistent upticks, the Clear Creek trend seems more randomized with more shifts in concavity and on a smaller scale.

## Trend Analysis
Research question: Has there been a monotonic trend in discharge in Clear Creek over the period of study?

9. Run a Seasonal Mann-Kendall test on the monthly discharge data. Inspect the overall trend and the monthly trends. 

```{r}
CCDischarge_timeseries <- ts(CCDischarge_monthly$Discharge_acftmo, frequency = 12,
                           start = c(1975, 1, 1), end = c(2021, 12, 31))
# Run SMK test
CCDischarge_trend <- smk.test(CCDischarge_timeseries)

# Inspect results
CCDischarge_trend
summary(CCDischarge_trend)
```

10. Is there an overall monotonic trend in discharge over time? Are there monthly trends over time? If so, are they positive or negative?

> There seems to be no significant trends over time or on the monthly scale. They seem to be slightly positive in the spring-summer months and negative for the rest of the year, with the closest to significant trend being in the middle of this potential wet season.


## Forecasting
Research question: can we predict discharge in Clear Creek moving into the future?

11. Run the auto.arima function on the Clear Creek time series to search for the best fit. Create an object that defines the best fit model. 
12. Make a prediction into the future and plot the future predictions. 

```{r}
auto.arima(CCDischarge_timeseries, trace = TRUE)

fit <- arima(CCDischarge_timeseries, c(1,0,0),seasonal = list(order = c(1,1,0), period = 12))

CCprediction <- predict(fit, n.ahead = 10*12)


ts.plot(CCDischarge_timeseries, CCprediction$pred, lty = c(1, 3))
```

13. How did the forecasting for Clear Creek compare to the Neuse River?

> It was much better! While it fails to take into account weather events that likely push discharge over the baseline, it does a good job showing the regular ebb and flow.