---
title: "Assignment 9: Mapping"
author: "Sasha Keller"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on mapping

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, check your PDF against the key and then submit your assignment completion survey at https://forms.gle/NDWEUu73LooFJPVM8 

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-04-12

## Setup

1. Load the tidyverse, LAGOSNE, maps, dataRetrieval, sf, and nhdplusTools packages. Set your ggplot theme (can be theme_classic or something else).


```{r setup}
library(tidyverse)
library(LAGOSNE)
library(maps)
library(dataRetrieval)
#install.packages("sf")
library(sf)
#install.packages("nhdplusTools")
library(nhdplusTools)

theme_set(theme_dark())
```

## LAGOS-NE

2. Choose five lakes in the LAGOS-NE database that are located within the same state and have chlorophyll data. Subset your data accordingly, and create two plots:

* A map of the five lakes within the state boundary, with each lake point as its own color.
* A boxplot with jittered points layered on top of chlorophyll concentrations in each lake (chlorophyll on y axis, lake on x axis), with each lake having a fill and/or color that matches the map. 

```{r}
# generate a map of U.S. states
states <- st_as_sf(map(database = "state", plot = TRUE, fill = TRUE, col = "white"))

# filter only states that are included in the LAGOSNE database
states.subset <- filter(states, ID=="new york")

# visualize state plot
ggplot(states.subset) +
  geom_sf(fill = "white")

# load LAGOSNE data frames
LAGOSdata <- lagosne_load()
LAGOSlocus <- LAGOSdata$locus
LAGOSstate <- LAGOSdata$state
LAGOSnutrient <- LAGOSdata$epi_nutr
LAGOSlimno <- LAGOSdata$lakes_limno


# Create a data frame to visualize chlorophyll a concentration
LAGOScombined <- 
  left_join(LAGOSnutrient, LAGOSlocus) %>%
  left_join(., LAGOSlimno) %>%
  left_join(., LAGOSstate) %>%
  filter(!is.na(state)) %>%
  select(lagoslakeid, sampledate, chla, lake_area_ha, maxdepth, nhd_lat, nhd_long, state)


chla.ny.summary <- LAGOScombined %>%
  filter(state=="NY")%>%
  group_by(lagoslakeid) %>%
  filter(lagoslakeid %in% c(122514,70859,51269,140000,9298))%>%
  summarise(chla.mean = mean(chla),
            area = mean(lake_area_ha),
            depth = mean(maxdepth),
            lat = mean(nhd_lat), 
            long = mean(nhd_long)) %>%
  drop_na()

#make spatial objects and maps
chla.spatial <- st_as_sf(chla.ny.summary, coords = c("long", "lat"), crs = 4326)



ggplot() +
  geom_sf(data = states.subset, fill = "white") +
  geom_sf(data = chla.spatial, aes(color = lagoslakeid), 
          alpha = 0.5, size = 4) #+
  #scale_color_viridis_c(trans = "log10", end = 0.8, direction = -1) +
  #labs(color = "Average Chlorophyll A Concentration (micrograms/L)") +
  #theme(legend.position = "top")

chlaboxplot<- ggplot(LAGOScombined,aes(x=as.factor(lagoslakeid), y=chla, fill=as.factor(lagoslakeid)))+
  geom_boxplot()+
  geom_jitter(alpha=0.5)
chlaboxplot
```


## NHDPlus

3. Delineate and map the watershed and flowpaths in the network upstream of the Eno River gage in Durham (USGS-02085070). Your map should include all stream orders.


```{r}
NeuseParams <- whatNWISdata(siteNumbers = "02085070")
NeuseCoords <- NeuseParams %>%
  select(site_no, dec_lat_va, dec_long_va) %>%
  distinct()

# Define the gage site as the starting point
start_point <- st_sfc(st_point(c(NeuseCoords$dec_long_va, NeuseCoords$dec_lat_va)), 
                      crs = 4269) # NAD83, commonly  used by US agencies
start_comid <- discover_nhdplus_id(start_point)
# start_point2 <- st_as_sf(data.frame(x = NeuseCoords$dec_long_va, y =  NeuseCoords$dec_lat_va), 
#                             coords = c("x", "y"), crs = 4269)

# Navigate the NLDI network
NLDI <- navigate_nldi(list(featureSource = "comid", featureID = start_comid), 
                          mode = "upstreamTributaries", 
                          distance_km = 1000)

# Extract watershed and flowpath information
subset_file <- tempfile(fileext = ".gpkg")
subset <- subset_nhdplus(comids = as.integer(NLDI$UT$nhdplus_comid),
                         output_file = subset_file,
                         nhdplus_data = "download", 
                         flowline_only = FALSE,
                         return_data = TRUE, overwrite = TRUE)

# Create data frames
flowline <- subset$NHDFlowline_Network
catchment <- subset$CatchmentSP
waterbody <- subset$NHDWaterbody

class(flowline)
class(catchment)
class(waterbody)

# find gages near watershed
gages <- get_nwis(AOI = catchment)
class(gages)

# find gages only within watershed
gages <- st_intersection(gages, catchment)


plot_nhdplus("USGS-02085070")

states <- st_as_sf(map(database = "state", plot = FALSE, fill = TRUE, col = "white"))
nc <- filter(states, ID == "north carolina")

ggplot(nc) +
  geom_sf(fill = "white") +
  geom_sf(data = flowline, aes(color = streamorde)) +
  labs(color = "Stream Order") +
  theme(legend.position = "top")

ggplot(catchment) +
  geom_sf(fill = "white", color = "gray", lwd = 0.5) +
  geom_sf(data = flowline, aes(color = streamorde)) +
  geom_sf(data = gages, color = "darkred", size = 1) +
  labs(color = "Stream Order") +
  theme(legend.position = "top")

max(flowline$totdasqkm)
summary(as.factor(flowline$streamorde))
gages<-get_nwis(AOI=catchment)
gages<-st_intersection(gages,catchment)
gages
```

4. What is the upstream drainage area for this gage, and how are the flowpaths distributed with regard to stream order?

> Upstream drainage spans 368 square kilometers, with 123 first order streams, 45 second order streams, 19 third order streams, and 28 fourth order streams.

5. Are there any other gage sites in the upstream catchment? If so, where are they?
> There are two additional gages: 02085039 and 02085000.
